---
title: "individual2"
author: "Colin Willenborg"
date: "9/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
dat <- readxl::read_xls('./GSS.xls')
dat <- dat %>%
  rename(
    Year = 'Gss year for this respondent',
    Happiness = 'General happiness',
    Religion = 'Religion in which raised',
    Political.Party = 'Political party affiliation',
    Sex = 'Respondents sex',
    Degree = 'Rs highest degree',
    Year.School = 'Highest year of school completed',
    Children = 'Number of children',
    Marital.Status = 'Marital status',
    Self.Emp = 'R self-emp or works for somebody',
    ID = 'Respondent id number',
    Ballot = 'Ballot used for interview'
  )
dat$Children <- as.numeric(dat$Children)
dat$Year.School <- as.numeric(dat$Year.School)
```

## Question 1: Education vs Political Party

```{r}
education <- select(dat, Year.School, Political.Party)
p <- ggplot(education, aes(x=Political.Party, y=Year.School, fill=Political.Party)) + geom_boxplot() + ggtitle("Education vs Political Party")
p + theme(axis.text.x=element_text(angle=90))
```

The above graph splits the respondents into groups based on Political Party and then shows a boxplot representation for the individual parties. People who identify as independent near democrat/republican on average have higher education than normal independent. The not strong republican and strong republican have a greater average education than the counter parts of not strong democrat and strong democrat. It is also interesting to note that a respondent identifying with a political party has a higher average education level than respondents with no education which could mean that the less educated you are the less you participate in politics.

## Question 2: Dataframe for Education above the median in Political Parties

```{r}
education <- dat %>% 
  select(Political.Party, Year.School) %>% 
  drop_na %>%
  group_by(Political.Party) %>%
  mutate(Year.School = as.numeric(Year.School, na.rm=TRUE)) %>%
  filter( Year.School >= median(Year.School, na.rm=TRUE))
str(education)
```
## Question 3: Happiness vs Marital Status

```{r}
marriedData <- dat %>% select(Happiness, Marital.Status) %>% drop_na
marriedData <- marriedData %>% filter(Marital.Status != 'No Answer' &
                                        Marital.Status != 'No answer' &
                                        Happiness != "Don't know" &
                                        Happiness != "No answer" &
                                        Happiness != 'Not applicable')
str(marriedData)
marriedData <- marriedData %>% mutate(Marital.Status = factor(Marital.Status, levels = c('Never married', 'Divorced', 'Seperated', 'Widowed', 'Married'))) %>%
  arrange(Marital.Status)
summary(marriedData)
ggplot(marriedData, aes(x=Marital.Status, fill=Happiness)) + geom_bar(position = 'fill') + ggtitle("Marital Status vs Happiness")
```
The graph shows that respondents that were married had a greater portion, by a significant margin, that wer very happy versus other groups.

## Question 4: Party Affiliation vs Happiness/Marital Status

```{r}
marriedData <- dat %>% select(Happiness, Political.Party, Marital.Status) %>% drop_na
marriedData <- marriedData %>% filter(Marital.Status != 'No Answer' &
                                        Marital.Status != 'No answer' &
                                        Happiness != "Don't know" &
                                        Happiness != "No answer" &
                                        Happiness != 'Not applicable' & 
                                        Political.Party != "Don't know" &
                                        Political.Party != "No answer" &
                                        Political.Party != 'Not applicable')
str(marriedData)
marriedData <- marriedData %>% mutate(Marital.Status = factor(Marital.Status, levels = c('Never married', 'Divorced', 'Seperated', 'Widowed', 'Married'))) %>%
  arrange(Marital.Status)
summary(marriedData)
ggplot(marriedData, aes(x=Marital.Status, fill=Happiness)) + geom_bar(position = 'fill') + facet_wrap(~marriedData$Political.Party) + theme(axis.text.x=element_text(angle=90))+ ggtitle("Political Party vs Happiness and Marital Status")
```
The trends for political party and marriage are fairly consistant throughout the graphs so it doesn't show any evidence that says political party has anything to do with marriage. 

## Question 5: Marital Status vs Education


```{r}
marriedData <- dat %>% select(Year.School, Marital.Status) %>% drop_na
marriedData <- marriedData %>% filter(Marital.Status != 'No Answer' &
                                        Marital.Status != 'No answer') %>%
  group_by(Marital.Status)
marriedData %>% 
  summarise(average = mean(Year.School))
```
The mean education level for each group of marital status is shown above and the only real difference is the widowed group has around 2 year average lower than the other groups. One reason for this could be the older respondents were more likely to be widowed and the education level of people has gone up in recent years so the older generation didn't have as much of a need to go on past high school. With out further evidence it is hard to make any predictions or assumptions.

```{r}
marriedData <- marriedData %>% mutate(Marital.Status = factor(Marital.Status, levels = c('Never married', 'Divorced', 'Seperated', 'Widowed', 'Married'))) %>%
  arrange(Marital.Status)
summary(marriedData)
ggplot(marriedData, aes(x=Marital.Status, y=Year.School)) + geom_boxplot() + ggtitle("Education vs Marital Status")
```
The graph above shows the relatively drastic difference when looking at widowed respondents versus the other three demographics. 

## Question 6: Effect of Children

```{r}
childrenHappiness <- dat %>% select(Happiness, Children) %>% 
  drop_na %>%
  filter(Happiness != "Don't know" & Happiness != "No answer" & 
           Happiness != 'Not applicable') %>%
  mutate(Children = factor(Children, levels = c(0, 1, 2, 3, 4, 5, 6, 7))) %>%
  arrange(Children) 
ggplot(childrenHappiness, aes(x=Children, fill=Happiness)) + geom_bar(position = 'fill') + ggtitle("Children vs Happiness")
```
First step I took was looking at the number of children vs Happiness but that turned up nothing remarkable like I was expecting. My hypothesis was that their would be a number that had more happiness to be considered the "ideal" number of children but that was not the case. It is all fairly even. So next I looked at the religoin of the respondent versus children to see if there was trend in the amount of kids per religion.

```{r}
temp <- dat %>% select(Religion) %>% unique
religionLevels <- c('Protestant', 'Catholic', 'Other', 'Jewish', 'Christian', 'Inter-nondenominational', 'Hinduism', 'Buddhism', 'Orthodox-christian', 'Moslem/islam', 'Other eastern', 'Native american')
childrenHappiness <- dat %>% select(Religion, Children) %>% 
  drop_na %>%
  filter(Religion != "Don't know" & Religion != "No answer" & 
           Religion != 'Not applicable') %>%
  mutate(Religion = factor(Religion, levels = religionLevels)) %>%
  arrange(Religion) 
ggplot(childrenHappiness, aes(x=Religion, y=Children)) + geom_boxplot() + theme(axis.text.x=element_text(angle=90))+ ggtitle("#Children vs Religoin")
```
The above graph shows that there is some differences with Catholics, Protestants, and Jewish respondents had the highest average and the middle chunk of religions, Christian, Inter-nondenominational, Hinduism, Buddhism and Orthodox-christian all were relatively the same. Protestant was the only, large religion, that had the range from the first quartile to third quartile not resting on 0 for number of children.


##Analysis: This lab was really well done by colin 
