---
title: "indivdual3"
author: "Allan Juarez"
date: "9/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
dat <- readxl::read_xls('C:/Users/Allan Juarez/OneDrive/Documents/DS202/ds202_lab3/GSS.xls')
summary(dat)
library(dplyr)
library(ggplot2)
library(classdata)
```

```{r}
education <- dat[c('Political party affiliation','Highest year of school completed')]  

education <- filter(education, education$'Highest year of school completed' != 'No answer')
education <- filter(education, education$'Highest year of school completed' != "Don't know")
education$'Highest year of school completed' <- as.numeric(education$'Highest year of school completed')

education <- aggregate(education$'Highest year of school completed',list(education$'Political party affiliation'), FUN = mean)
education <- education %>% rename(party = Group.1, years = x)
 
ggplot(education, aes(x=party, y=years)) + geom_bar(stat = "identity")+coord_flip()
```

1. Looking at the average years of school for every party. YOu can see that there are about three types of the two parties. Strong, Not Str, INd, in every category republicans average more years of school than democrats. The lowest average of years in school tended to not know their affiliation.

```{r}
second <-dat

education <- dat[c('Political party affiliation','Highest year of school completed')] 
education <- filter(education, education$'Highest year of school completed' != 'No answer')
education <- filter(education, education$'Highest year of school completed' != "Don't know")
education$'Highest year of school completed' <- as.numeric(education$'Highest year of school completed') 
education <- aggregate(education$'Highest year of school completed',list(education$'Political party affiliation'), FUN = median)
education
second <- filter(second, second$'Highest year of school completed' != 'No answer')
second <- filter(second, second$'Highest year of school completed' != "Don't know")

second$'Highest year of school completed' <- as.numeric(second$'Highest year of school completed')
strongRep <- filter(second,second$'Political party affiliation' == 'Strong republican' & second$'Highest year of school completed' >13)

strongDem <- filter(second,second$'Political party affiliation' == 'Strong democrat' & second$'Highest year of school completed' >12)

other <- filter(second,second$'Political party affiliation' == 'Other party' & second$'Highest year of school completed' >13)

notStrRep <- filter(second,second$'Political party affiliation' == 'Not str republican' & second$'Highest year of school completed' >13)

notStrDem <- filter(second,second$'Political party affiliation' == 'Not str democrat' & second$'Highest year of school completed' >12)

no <- filter(second,second$'Political party affiliation' == 'No answer' & second$'Highest year of school completed' >12)

independent <- filter(second,second$'Political party affiliation' == 'Independent' & second$'Highest year of school completed' >12)
inrep <- filter(second,second$'Political party affiliation' == 'Ind,near rep' & second$'Highest year of school completed' >13)

indem <- filter(second,second$'Political party affiliation' == 'Ind,near dem' & second$'Highest year of school completed' >13)

know <- filter(second,second$'Political party affiliation' == "Don't know" & second$'Highest year of school completed' >11)

betterEduacatied <- rbind(strongRep,strongDem)
betterEduacatied <- rbind(betterEduacatied,other)
betterEduacatied <- rbind(betterEduacatied,notStrRep)
betterEduacatied <- rbind(betterEduacatied,notStrDem)
betterEduacatied <- rbind(betterEduacatied,independent)
betterEduacatied <- rbind(betterEduacatied,inrep)
betterEduacatied <- rbind(betterEduacatied,indem)
betterEduacatied <- rbind(betterEduacatied,know)

betterEduacatied

```
2. the median years were used to figure out years
```{r}
marriage <- dat[c('General happiness','Marital status','Political party affiliation')]
marriage <- filter(marriage, marriage$'General happiness' == 'Very happy' | marriage$'General happiness' == 'Pretty happy' | marriage$'General happiness' == 'Not too happy')
marriage <-
marriage <- table(marriage)
marriage <- as.data.frame(marriage)

ggplot(marriage, aes(x=Marital.status, y=Freq)) + geom_bar(aes(fill = General.happiness),stat = "identity",position = "dodge")+coord_flip()
```
3. By the look of the visual and numerically. THe overall person is happy no matter their martial status, but Looking at the married group there are a lot more people that are happy than any other group. In the numerical representation you can see more people are happy across all group, but visually you can see that a lot of people responded to being married and that they were happy. 

```{r}

partyMarriage <- dat[c('General happiness','Marital status','Political party affiliation')]
partyMarriage <- filter(partyMarriage, partyMarriage$'General happiness' == 'Very happy' | partyMarriage$'General happiness' == 'Pretty happy' | partyMarriage$'General happiness' == 'Not too happy') 

partyMarriage <- filter(partyMarriage, partyMarriage$'Political party affiliation' != "Don't know" | partyMarriage$'Political party affiliation' != "No answer")

ggplot(partyMarriage, aes(partyMarriage$'Marital status', fill =partyMarriage$'General happiness' ))+ geom_bar(position = "dodge")+ facet_wrap(~partyMarriage$'Political party affiliation') +coord_flip()
```
4. Looking at this there doesnt seem to be a trend between martial status and political party they are all very simlar.

```{r}
happy <- dat[c('Marital status','Highest year of school completed')]  

happy <- filter(happy, happy$'Highest year of school completed' != 'No answer')
happy <- filter(happy, happy$'Highest year of school completed' != "Don't know")
happy <- filter(happy, happy$'Marital status' != "No answer")
happy$'Highest year of school completed' <- as.numeric(happy$'Highest year of school completed')
happy <- aggregate(happy$'Highest year of school completed',list(happy$'Marital status'), FUN = mean)
happy <- happy %>% rename(party = Group.1, years = x)
happy
 
ggplot(happy, aes(x=party, y=years)) + geom_bar(stat = "identity")+coord_flip()
```
5. all the statuses are pretty similar id say there is at least a visual difference on widowed and separated apart from the rest of the groups. Numerically widowed is the lowest with 11 and divorced, married and never is hovering around 13.

```{r}
employed <-dat[c('R self-emp or works for somebody','General happiness')]   
employed <- filter(employed, employed$'General happiness' == 'Very happy' | employed$'General happiness' == 'Pretty happy' | employed$'General happiness' == 'Not too happy')
employed <- filter(employed, employed$'R self-emp or works for somebody' != 'No answer' )
employed <- filter(employed, employed$'R self-emp or works for somebody' != 'Not applicable')
 employed <- filter(employed,employed$'R self-emp or works for somebody' != "Don't know" )
employed <- table(employed)
employed <- as.data.frame(employed)
employed
ggplot(employed, aes(x=R.self.emp.or.works.for.somebody, y=Freq)) + geom_bar(aes(fill = General.happiness),stat = "identity",position = "dodge")+coord_flip()
```
6. For individual we compared the happiness on the person and checked if they were unemployed or not. There didnt really seem to be any trend between them