---
title: "Team"
author: "Colin Willenborg, Allan Juarez, Adam Brandt"
date: "9/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
dat <- readxl::read_xls('./GSS.xls')
dat <- dat %>%
  rename(
    Year = 'Gss year for this respondent',
    Happiness = 'General happiness',
    Religion = 'Religion in which raised',
    Political.Party = 'Political party affiliation',
    Sex = 'Respondents sex',
    Degree = 'Rs highest degree',
    Year.School = 'Highest year of school completed',
    Children = 'Number of children',
    Marital.Status = 'Marital status',
    Self.Emp = 'R self-emp or works for somebody',
    ID = 'Respondent id number',
    Ballot = 'Ballot used for interview'
  )
dat$Children <- as.numeric(dat$Children)
dat$Year.School <- as.numeric(dat$Year.School)
```

## Question 1

## Question 2

## Question 3

## Question 4

## Question 5

## Question 6: Children, Religion and Self-employed

## Children

```{r}
childrenHappiness <- dat %>% select(Happiness, Children) %>% 
  drop_na %>%
  filter(Happiness != "Don't know" & Happiness != "No answer" & 
           Happiness != 'Not applicable') %>%
  mutate(Children = factor(Children, levels = c(0, 1, 2, 3, 4, 5, 6, 7))) %>%
  arrange(Children) 
ggplot(childrenHappiness, aes(x=Children, fill=Happiness)) + geom_bar(position = 'fill') + ggtitle("Children vs Happiness")
```
First step I took was looking at the number of children vs Happiness but that turned up nothing remarkable like I was expecting. My hypothesis was that their would be a number that had more happiness to be considered the "ideal" number of children but that was not the case. It is all fairly even. So next I looked at the religoin of the respondent versus children to see if there was trend in the amount of kids per religion.

```{r}
temp <- dat %>% select(Religion) %>% unique
religionLevels <- c('Protestant', 'Catholic', 'Other', 'Jewish', 'Christian', 'Inter-nondenominational', 'Hinduism', 'Buddhism', 'Orthodox-christian', 'Moslem/islam', 'Other eastern', 'Native american')
childrenHappiness <- dat %>% select(Religion, Children) %>% 
  drop_na %>%
  filter(Religion != "Don't know" & Religion != "No answer" & 
           Religion != 'Not applicable') %>%
  mutate(Religion = factor(Religion, levels = religionLevels)) %>%
  arrange(Religion) 
ggplot(childrenHappiness, aes(x=Religion, y=Children)) + geom_boxplot() + theme(axis.text.x=element_text(angle=90))+ ggtitle("#Children vs Religoin")
```
The above graph shows that there is some differences with Catholics, Protestants, and Jewish respondents had the highest average and the middle chunk of religions, Christian, Inter-nondenominational, Hinduism, Buddhism and Orthodox-christian all were relatively the same. Protestant was the only, large religion, that had the range from the first quartile to third quartile not resting on 0 for number of children.

## Religion

## Self-Employed


